\documentclass[12pt]{paper} 
\usepackage{amsmath, amsfonts, amsthm} % Math packages
\usepackage{listings} % Code listings, with syntax highlighting
\usepackage[english]{babel} % English language hyphenation
\usepackage{graphicx} % Required for inserting images
\usepackage{booktabs} % Required for better horizontal rules in tables
\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs

\usepackage{enumitem} % Required for list customisation
\setlist{noitemsep} % No spacing between list items

%----------------------------------------------------------------------------------------
%	DOCUMENT MARGINS
%----------------------------------------------------------------------------------------

\usepackage{geometry} % Required for adjusting page dimensions and margins

\geometry{
	paper=letterpaper, % Paper size, change to letterpaper for US letter size
	top=2.5cm, % Top margin
	bottom=3cm, % Bottom margin
	left=3cm, % Left margin
	right=3cm, % Right margin
	headheight=0.75cm, % Header height
	footskip=1.5cm, % Space from the bottom margin to the baseline of the footer
	headsep=0.75cm, % Space from the top margin to the baseline of the header
	%showframe, % Uncomment to show how the type block is set on the page
}

%----------------------------------------------------------------------------------------
%	FONTS
%----------------------------------------------------------------------------------------

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Use 8-bit encoding

\usepackage{fourier} % Use the Adobe Utopia font for the document

%----------------------------------------------------------------------------------------
%	HEADERS AND FOOTERS
%----------------------------------------------------------------------------------------

\usepackage{scrlayer-scrpage} % Required for customising headers and footers

\ohead*{} % Right header
\ihead*{} % Left header
\chead*{} % Centre header

\ofoot*{} % Right footer
\ifoot*{} % Left footer
\cfoot*{\pagemark} % Centre footer

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\title{	
Problem Set 4--STAT 7200
}

\author{} % Your name
\date{} % Today's date (\today) or a custom date

\begin{document}

\maketitle % Print the title

%----------------------------------------------------------------------------------------

Unless otherwise stated, the significance level is $\alpha = 0.05$.

\begin{enumerate}
	\item\label{prob:differences} Let $(\mathbf{Y}_1, \mathbf{X}_1), \ldots, (\mathbf{Y}_n, \mathbf{X}_n)$ be a paired random sample. Construct a test statistic for testing the equality of the means, i.e.\ $H_0: E(\mathbf{Y}_i) = E(\mathbf{X}_i)$. Give the distribution of your test statistic under the null hypothesis, and describe the distributional assumptions you are making.
	
	\vspace{1.5cm}\item\label{prob:gap} Using the result of Problem~\ref{prob:differences} on the \texttt{gapminder} dataset, test the null hypothesis of no change in infant mortality, life expectancy and fertility from 2012 to 2013 (only keep countries with complete observations for both years). Construct the simultaneous confidence intervals and the Bonferroni-corrected univariate confidence intervals. Discuss the results. 
	
	\vspace{1.5cm}\item Investigate the multivariate normality assumption for the dataset in Problem~\ref{prob:gap}. Perform a permutation test and compare the results with what you obtained earlier.
	
	\vspace{1.5cm}\item\label{prob:contrasts} Let $\mathbf{Y}_1, \ldots, \mathbf{Y}_n \sim N_p(\mu, \Sigma)$, with $\Sigma$ positive definite, and let $\bar{\mathbf{Y}}$ and $S_n$ be the sample mean and covariance, respectively. Let $C$ be a $(p-1)\times p$ matrix of rank $p-1$ such that $C\mathbf{1} = 0$, i.e. the rows of $C$ sum to zero. Prove the following:
	\begin{enumerate}
		\item The following null hypotheses are equivalent:
		$$H_0: C\mu = 0 \Leftrightarrow H_0: \mu_1 = \cdots = \mu_p.$$
		\item Any $p-1$ columns of $C$ are linearly independent, which implies that we can write
		$$ C = A (I_{p-1}, -\mathbf{1}),$$
		for some invertible $(p-1)\times(p-1)$ matrix $A$.
		\item Using part (b), show that the test statistic
		$$ T_2 = n(C\bar{\mathbf{Y}})^T(CS_n C^T)^{-1}(C\bar{\mathbf{Y}})$$
		does not depend on $C$.
		\item Under the null hypothesis of part (a), the distribution of $T^2$ is
		$$ T^2 \sim \frac{(n - 1)(p - 1)}{n - p + 1}F_\alpha(p-1, n - p +1).$$
	\end{enumerate}

	\vspace{1.5cm}\item Using the result of Problem~\ref{prob:contrasts} on the Ramus dataset, perform a hypothesis test that the ramus lengths are constant over time. Construct the simultaneous confidence intervals and the Bonferroni-corrected univariate confidence intervals. Discuss the results. 
	
	\vspace{1.5cm}\item Let $GL_p$ be the set of invertible $p\times p$ matrices.\footnote{This set is actually called the \emph{general linear group}, and it is ubiquitous in mathematics.} Any matrix $A\in GL_p$ induces a transformation on the data $\mathbf{Y}_i \mapsto A\mathbf{Y}_i$, which also induces a transformation on the sample statistics
	$$ (\bar{\mathbf{Y}}, S_n) \mapsto (A\bar{\mathbf{Y}}, AS_n A^T),$$
	and on the population parameter
	$$ (\bar\mu, \Sigma) \mapsto (A\mu, A\Sigma A^T).$$
	We say a test statistic $f(\bar{\mathbf{Y}}, S_n)$ is \emph{invariant} if it yields the same value on the transformed data:
	$$ f(\bar{\mathbf{Y}}, S_n) = f(A\bar{\mathbf{Y}}, AS_n A^T), \qquad\mbox{for all } A\in GL_p.$$
	By choosing an appropriate $A$, show that an invariant test statistic satisfies
	$$ f(\bar{\mathbf{Y}}, S_n) = f((\bar{\mathbf{Y}}^T S_n^{-1}\bar{\mathbf{Y}})^{1/2}, I).$$
	Conclude that an invariant test statistic depends on the data only through the statistic $T^2 = n\bar{\mathbf{Y}}^T S_n^{-1}\bar{\mathbf{Y}}$.
\end{enumerate}

\end{document}
